\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}

### 2016-12-28 latex/article_vol2.tex
First bunch of commits

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}
\usepackage[nogin]{Sweave}
\usepackage{pdfpages}
\usepackage{xspace}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}

@@

@@

By default, a pdf overview document is produced, saved to the disc (in
the working directory) and opened for immediate inspection. Turning to
Figure~\ref{fig:example1}, we see that such a data cleaning output
document consists of two parts. First, an overview of what was done is
presented under the title \textit{Data cleaning summary}. Secondly,
each variable in the dataset is presented in turn using (up to) three
tools in the \textit{Variable list}: A table summarizing key features
of the variable, a figure visualizing its distribution and potentially
also a list of flagged issues. For instance, in the \R{numeric}-type
variable \R{var2} from \R{toyData}, \R{clean()} has identified two
values that are suspected to be miscoded missing values (\R{999} and
\R{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully. 
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

@@

\smallskip Control output and post-processing \\
\quad \R{output} & Type of output file to be produced (html, or pdf) & \R{"pdf"} \\
\quad \R{render} & Should the output file be rendered from markdown? & \R{TRUE} \\
\quad \R{openResult} & If a  pdf/html file is rendered, should it automatically open afterwards, and if not, should the \R{rmarkdown} file? & \R{TRUE} \\

@@

\begin{figure}[tb]
\begin{description}
\item[Input] A dataset
	\begin{itemize} 
		\item This should be of type \R{data.frame} or \R{tibble} \hl{or data.table?}
	\end{itemize}
\item[Create contents] For each variable in the dataset, do the following:
	\begin{description}
		\item[Stage 1:] Pre-checks 
			\begin{itemize}
				\item Is the variable suitable for summarization, visualization and checks?
				\begin{description}
					\item[Yes:] Go to stage 2. 
					\item[No:] Move on to the next variable.
				\end{description}
			\end{itemize}
		\item[Stage 2:] SVC-steps
			\begin{description}
				\item[Summarize] Call \R{summarize()} to produce a summary table describing the variable. What features enter this table depends on the data class of the variable.
				\item[Visualize]  Call \R{visualize()} to produce a plot visualizing the distribution of the variable. 
				\item[Check]  Call \R{check()} to apply quality- and error checks to the variable. What checks are used depends on the data class of the variable.
			\end{description}
	\end{description}
\item[Output] Files for a overview document are saved to the disc and possibly also opened.
	\begin{itemize}
		\item Always a \R{rmarkdown} (.Rmd) file
		\item Possible also a html or pdf file
	\end{itemize}
\end{description}
\caption{\hl{quite odd having a figure like this? maybe do a flowchart?}}
\label{figure:cleanStructure}
\end{figure}

@@

For example, to get a summary document that only contains the
variables with potential problems, and with a limit of maximum 10
printed potential errors, we can write (output not shown)

\begin{Schunk}
\begin{Sinput}
> clean(toyData, onlyProblematic=TRUE, maxProbVals=10)
\end{Sinput}
\end{Schunk}

The final rendering of the generated

By default, \R{clean()} runs all the summary,
visualization, and check functions that are
implemented. Table~\ref{table.SVCfunctions} lists these functions but
we can also use the \R{allSummaryFunctions()},
\R{allVisualFunctions()}, and \R{allCheckFunctions()} function in
\proglang{R} to obtain a list. For example the implemented summary
functions are:

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\ 
  \hline \smallskip
  \textbf{\R{summaryFunction}s}  \smallskip \\
  \quad \R{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\ 
  \quad \R{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ &  \\ 
  \quad \R{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \smallskip \\
 \textbf{\R{visualFunction}s} \smallskip \\
  \quad \R{basicVisual} & Histograms and barplots using graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{standardVisual} & Histograms and barplots using ggplot2 &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \smallskip \\
 \textbf{\R{checkFunction}s} \smallskip \\
 \quad \R{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\ 
 \quad \R{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & \\ 
 \quad \R{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ &  \\ 
 \quad \R{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\ 
 \quad \R{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &   \\ 
 \quad \R{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &   \\ 
 \quad \R{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & \smallskip   \\ 
 \hline
\end{tabular}
\caption{\hl{Blabla, mention that C is character, F is factor, I is integer, L is labelled, B is logical (boolean), N is numeric and D is Date. \newline Also: Check that everything in here is correct (i.e. corresponds to the output of allSummaryFunctions() etc when makeXfunction has been fixed.}}
\label{table.SVCfunctions}
\end{table}

\hl{
\begin{itemize}
\item Introduce the relevant \R{clean}-arguments and the \R{defaultWhateverSummaries} etc.-functions
\item Introduce \R{allSummaryFunctions()} etc. and present a table corresponding to the output of this call
\item Small examples: 
\begin{itemize}
\item Add a function to one of the XXXSummaries/XXXVisuals/XXXChecks-arguments (still calling default options)
\item Remove all but a single function from one of these arguments
\item Describe what happens if the argument is NULL
\end{itemize}
\end{itemize}
}

@@

In this section, we discuss how to control every step in \R{clean()} that actually involves a function being called on variables from the dataset. There are two stages in which this occurs, as mentioned in Figure~\ref{figure:cleanStructure}, namely:
\begin{enumerate}
\item In the precheck functions
\item In the summarize/visualize/check (SVC) step
\end{enumerate}
Each of these stages are controllable using appropriate function parameters in \code{clean}. In the above, we presented the default \pkg{cleanR} settings and how to tweak them into providing a slightly different data cleaning outputs. However, if for instance the dataset at hand requires completely different visualizations, more control is needed. \pkg{cleanR} uses three different types of functions for performing all stages in the above, namely \code{summaryFunction}s, \code{visualFunction}s and \code{checkFunction}s. \hl{Something like "the avaialable options for functions used in the precheck- and check steps are obtainable by calling \R{allCheckFunction()}, blablabla, similarly with visualFunctions and summaryFunctions. Mention Section 4 and the fact that one can expand the possibilities of e.g. allVisualFunctions() by producing new functions quite easily.}

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()}, and
thus they can also be used here. For instance, if we only want the
result of the check for miscoded missing values, we write
\begin{Schunk}
\begin{Sinput}
> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}
The \R{numericChecks} argument XXX.. ... not defined

An equivalent way to call only a single, specific \R{checkFunction}
such as \R{identifyMissing} (see \ref{table.SVCfunctions} for a list
of check functions) is by using it directly on the variable, i.e.
\begin{Schunk}
\begin{Sinput}
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN. 
\end{Soutput}
\end{Schunk}
% $

@@

\hl{
\begin{itemize}
\item Do an example with visualize() and summarize(), like the one with check(). Especially visualize and to doEval = T thing needs a bit of special attention.
\item  Mention allCheckFunctions() etc. again here
\item Mention check(), visualize() and summarize() modes for data.frames. Maybe also advice against it, at it will often produce a lot of information at once, and such large amounts of information really should be documented.
\end{itemize}
}

@@

\begin{table}[ht]
\scriptsize
\begin{tabular}{p{0.2\linewidth}p{0.2\linewidth}p{0.2\linewidth}p{0.2\linewidth}}
& \R{summaryFunction} & \R{visualFunction} & \R{checkFunction} \\
\hline
Input (required) & \R{v} - a variable vector \newline \R{...}  &  \R{v} - a variable vector \newline \R{vnam} - the variable name (as character string) \newline \R{doEval} - a logical (\R{TRUE}/\R{FALSE}) controlling the output type of the function & \R{v} - a variable vector \newline \R{nMax} - an integer (or \R{Inf}), controlling how many problematic values are printed, if relevant \newline \R{...}  \\
Input (optional) &  \R{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \R{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g. a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries \newline \R{\$feature} - a label for the summary value (as character string) \newline \R{\$result} - the result of the summary (as character string) & A character string with \R{R} code for producing a plot. This code should be standalone, i.e. should include the data if necessary. & A list with entries \newline \R{\$problem} - a logical identifying whether an issue was found \newline \R{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \R{rmarkdown} \\
Output (recommended) & A \R{summaryResult} object (i.e. an attributed list with entries \R{\$feature}, \R{\$result} and \R{\$value}, the latter being the values from \R{\$result} in their original format). & \begin{tabular}{p{0.4\linewidth} | p{0.4\linewidth}} If \R{doEval} is \R{TRUE}: \newline A plot that will be opened by the graphic device in \R{R}. & If \R{doEval} is \R{FALSE}: \newline   A text string with \R{R} code, as described above. \end{tabular} & A \R{checkResult} object (an attributed list with entries \R{\$problem}, \R{\$message} and \R{\$problemValues}, the latter being either \R{NULL} or the problem causing values, as they were found in \R{v}, whichever is relevant. \\
Tools available for producing the function & \R{summaryResult()} & - & \R{messageGenerator()} \newline \R{checkResult()} 
\end{tabular}
\caption{\hl{blabla. Fix formatting! Maybe use multirow and raggedRight stuff?}}
\label{table.functionTypes}
\end{table}

Though the discussion in the above paints a picture of cleanR as a user-friendly package which requires practically no knowledge of \R{R}, one should not be mistaken to think that it is not customizable. In fact, the main function of \pkg{cleanR}, \code{clean}, is mainly a tool for formatting the results from various checking-, summary- and visualization functions. Thus, the actual work underlying a \pkg{cleanR} output file can be anything or nothing - depending on the arguments given to \code{clean}. Specfically, user made functions can be added to the SVC-function arguments discussed above, e.g. \R{factorSummaries}, \R{allVisuals} or \R{numericChecks}: All that is needed is to specify their names, just like the names of the build-in SVC functions are specified. However, not just any function can be called from these three steps, and therefore, we will now present how \R{summaryFunction}s, \R{visualFunction}s and \R{checkFunction}s are made. \hl{Mention something about the interactive mode here as well?} \\

>>>>

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}
\usepackage[nogin]{Sweave}
\usepackage{pdfpages}
\usepackage{xspace}
\usepackage{array}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}

@@

\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

By default, an \proglang{R} markdown file and rendered pdf overview
document is produced, saved to the disc (in the working directory) and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data cleaning output document consists of two
parts. First, an overview of what was done is presented under the
title \textit{Data cleaning summary}. Secondly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and potentially also a
list of flagged issues. For instance, in the \R{numeric}-type variable
\R{var2} from \R{toyData}, \R{clean()} has identified two values that
are suspected to be miscoded missing values (\R{999} and \R{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

@@

\smallskip Control output and post-processing \\
\quad \R{output} & Type of output file to be produced (html, or pdf) & \R{"pdf"} \\
\quad \R{render} & Should the output file be rendered from markdown? & \R{TRUE} \\
\quad \R{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \R{rmarkdown} file be opened? & \R{TRUE} \\
\quad \R{replace} & Overwrite an existing markdown file with the same name? & \R{FALSE} \\

@@

\begin{figure}[tb]
% \begin{description}
% \item[Input] A dataset
% 	\begin{itemize} 
% 		\item This should be of type \R{data.frame} or \R{tibble} \hl{or data.table?}
% 	\end{itemize}
% \item[Create contents] For each variable in the dataset, do the following:
% 	\begin{description}
% 		\item[Stage 1:] Pre-checks 
% 			\begin{itemize}
% 				\item Is the variable suitable for summarization, visualization and checks?
% 				\begin{description}
% 					\item[Yes:] Go to stage 2. 
% 					\item[No:] Move on to the next variable.
% 				\end{description}
% 			\end{itemize}
% 		\item[Stage 2:] SVC-steps
% 			\begin{description}
% 				\item[Summarize] Call \R{summarize()} to produce a summary table describing the variable. What features enter this table depends on the data class of the variable.
% 				\item[Visualize]  Call \R{visualize()} to produce a plot visualizing the distribution of the variable. 
% 				\item[Check]  Call \R{check()} to apply quality- and error checks to the variable. What checks are used depends on the data class of the variable.
% 			\end{description}
% 	\end{description}
% \item[Output] Files for a overview document are saved to the disc and possibly also opened.
% 	\begin{itemize}
% 		\item Always a \R{rmarkdown} (.Rmd) file
% 		\item Possible also a html or pdf file
% 	\end{itemize}
% \end{description}

% Define block styles
\tikzstyle{decision} = [diamond, draw, fill=blue!20, 
    text width=5.5em, text badly centered, node distance=3cm, inner
    sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
    text width=6em, text centered, rounded corners, minimum
    height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
    text width=5em,
    minimum height=2em]

\begin{center}    
\begin{tikzpicture}[node distance = 2cm, auto,thick,scale=0.75, every node/.style={transform shape}]
    % Place nodes
    \node [block] (init) {Get next variable and run pre checks};
    \node [cloud, above of=init] (input) {Input \texttt{data.frame} or \texttt{tibble}};
%    \node [cloud, right of=init] (system) {system};
    \node [decision, right of=init, node distance=4cm] (precheck) {Is variable suitable for inclusion};
    \node [block, right of=precheck, node distance=4cm] (summarize)
    {Run \texttt{summarize()} to produce summary table};
    \node [block, below of=summarize, node distance=3cm] (visualize)
    {Run \texttt{visualize()} to plot variable};
    \node [block, below of=visualize, node distance=2.5cm] (check)
    {Call \texttt{check()} to run error checks};
    \node [decision, below of=check, node distance=2.7cm] (done) {More
      variables?};
    \node [block, right of=done, node distance=3.5cm] (stop) {Write
      \proglang{R} markdown file};
    \node [cloud, below of=stop, node distance=3.5cm] (render) {Render
      markdown and possiby open};
    % Draw edges
    \path [line] (summarize) -- (visualize);
    \path [line] (visualize) -- (check);
    \path [line] (check) -- (done);
    \path [line] (done.south) -- +(0,-10pt) -| node [near start] {yes} (init);
%    \path [line] (identify) -- (evaluate);
%    \path [line] (evaluate) -- (decide);
%    \path [line] (init) -| node [near start] {yes} (precheck);
    \path [line] (init) -- (precheck);
    \path [line] (precheck) -- node [near start] {yes} (summarize);
    \path [line] (precheck) |- node [near start] {no} (done);
    \path [line] (done) -- node [near start] {no} (stop);
%    \path [line] (update) |- (identify);
%    \path [line] (decide) -- node {no}(stop);
    \path [line,dashed] (input) -- (init);
%    \path [line,dashed] (system) -- (init);
%    \path [line,dashed] (system) |- (evaluate);
\end{tikzpicture}
\end{center}

\caption{Overall illustration of the stages undertaken when running
  \R{clean()}. Each variable is checked for eligibility before
  running \R{summarize()}, \R{visualize()}, and \R{check()}, and the
  resulting \proglang{R} markdown file may be rendered and opened.}
\label{figure:cleanStructure}
\end{figure}

@@

For example, to get a summary document in html format that only
contains the variables with potential problems, and with a limit of
maximum 10 printed potential errors for each variable we can write
(output not shown). Also, we can add the argument \R{replace=TRUE} in
order to force \R{clean()} to overwrite any existing files produced by
\R{clean()}.

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html", onlyProblematic=TRUE, maxProbVals=10,
        replace=TRUE)
\end{Sinput}
\end{Schunk}

The final rendering of the generated markdown file is controlled with
the \R{render} and \R{openResult} arguments which both default to
\R{TRUE} \hl{Bliver det ved med at være sandt?}. \R{render} determines
if the \proglang{R} markdown file produced should be rendered using
the \pkg{rmarkdown} package and \R{openResult} decides whether the
output html or pdf file should be opened. The following command
produces an \proglang{R} markdown file containing the data cleaning
process information but the markdown file is not rendered into html
and it is not opened.

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html", render=FALSE, openResult=FALSE)
\end{Sinput}
\end{Schunk}

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\ 
  \hline \smallskip
  \textbf{\R{summaryFunction}s}  \smallskip \\
  \quad \R{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\ 
  \quad \R{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\ 
  \quad \R{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \smallskip \\
 \textbf{\R{visualFunction}s} \smallskip \\
  \quad \R{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{standardVisual} & Histograms and barplots using ggplot2 &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \smallskip \\
 \textbf{\R{checkFunction}s} \smallskip \\
 \quad \R{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\ 
 \quad \R{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\ 
 \quad \R{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\ 
 \quad \R{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\ 
 \quad \R{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\ 
 \quad \R{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
 \quad \R{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\ 
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{cleanR}. The variable
  classes C, F, I, L, B, N, and D, refer to \R{character}, \R{factor},
  \R{integer}, \R{labelled}, \R{logical} (boolean), \R{numeric}, and \R{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

\pkg{cleanR} uses three
different types of functions for performing all steps in the above,
namely \code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.
By default, \R{clean()} runs all the summary, visualization, and check
functions currently implemented for each variable type as part of the
algorithm (see
Figure~\ref{figure:cleanStructure}). Table~\ref{table.SVCfunctions}
lists all these functions but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} function in \proglang{R} to print an overview
list. For example the implemented summary functions are:

@@

@@

Thus we can see, for example, that the summary table for a
\R{numeric}, \R{integer}, or \R{Date} variable will include
information about the minimum and maximum values, while all variable
classes will produce some kind of information about the central
tendency (i.e., mode or median).

We can control the default summaries and checks for each variable type
through the \R{XXXSummaries} and \R{XXXChecks} arguments where \R{XXX}
represents a variable type, e.g., \R{factorChecks} for factors,
\R{numericChecks} for numeric variables, etc. These arguments accept a
vector of summary or check functions that should be run for a
particular variable type and the default values for each of these
arguments (i.e., the functions from Table~\ref{table.SVCfunctions}
that are applied) can be obtained through the \R{defaultXXXChecks()}
and \R{defaultXXXSummaries()} functions, where \R{XXX} again refers to
the variable type.

For example, the default summaries being used for a factor variable is 
\begin{Schunk}
\begin{Sinput}
> defaultFactorSummaries()
\end{Sinput}
\begin{Soutput}
[1] "variableType" "countMissing" "uniqueValues" "centralValue"
\end{Soutput}
\end{Schunk}

We can change the summaries (and similarly the checks) by setting the
corresponding arguments when calling \R{clean()}. For example, to get
only the variable type and the central tendency listed in the summary
table of each variable we write

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html",  
        characterSummaries = c("variableType", "centralValue"),
        factorSummaries = c("variableType", "centralValue"),
        labelledSummaries = c("variableType", "centralValue"),
        numericSummaries = c("variableType", "centralValue"),
        integerSummaries = c("variableType", "centralValue"),
        logicalSummaries = c("variableType", "centralValue"),
        dateSummaries = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

In this particular case, where we specify the same set of summary
functions for each variable type, we can use the simpler argument
\R{allSummaries} which overrides the summary functions for all
variable types. Thus, the same result as above could be obtained with

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html",  
        allSummaries = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

Similarly, the checks applied are set with the \R{XXXChecks}
arguments. The default checks being applied to a factor is

\begin{Schunk}
\begin{Sinput}
> defaultFactorChecks()
\end{Sinput}
\begin{Soutput}
[1] "identifyMissing"    "identifyWhitespace" "identifyLoners"    
[4] "identifyCaseIssues" "identifyNums" 
\end{Soutput}
\end{Schunk}

Now, if we only wanted to apply the identify whitespace function for
factor variables then we would need to set the \R{factorChecks} accordingly

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html",  
        factorChecks = c("identifyWhitespace"))
\end{Sinput}
\end{Schunk}

or we could remove checks for factor altogether by setting the
corresponding argument to \R{NULL} in which case factor variables will
not be checked for any potential errors.

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html", factorChecks = NULL)
\end{Sinput}
\end{Schunk}

%\hl{
%\begin{itemize}
%\item Introduce the relevant \R{clean}-arguments and the \R{defaultWhateverSummaries} etc.-functions
%\item Introduce \R{allSummaryFunctions()} etc. and present a table corresponding to the output of this call
%\item Small examples: 
%\begin{itemize}
%\item Add a function to one of the XXXSummaries/XXXVisuals/XXXChecks-arguments (still calling default options)
%\item Remove all but a single function from one of these arguments
%\item Describe what happens if the argument is NULL
%\end{itemize}
%\end{itemize}
%}

As indicated in Figure~\ref{figure:cleanStructure} there are two stages
where \R{clean()} applies function(s) to each of the variables: 
\begin{enumerate}
\item In the precheck functions
\item As part of the summarize/visualize/check (SVC) steps
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \R{clean()}, and above we have shown examples of how to
tweak them to modify the data cleaning outputs. However, if for
instance the dataset at hand requires completely different
visualizations then more control is needed, and
Section~\ref{sec:extending} explains how to modify, and expand the
possibilities by producing new functions quite easily.

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\R{XXXChecks} argument (e.g., \R{numericChecks}, \R{factorChecks},
etc. as described in Section~\ref{sec:usingcleanR}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or the \R{allCheckFunctions()} provides
overviews of the available check functions.
\begin{Schunk}
\begin{Sinput}
> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

An equivalent way to call only a single, specific \R{checkFunction}
such as \R{identifyMissing} (see \ref{table.SVCfunctions} for a list
of check functions) is by using it directly on the variable, e.g.,
\begin{Schunk}
\begin{Sinput}
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN. 
\end{Soutput}
\end{Schunk}
% $

@@

Similarly, the \R{visualize()} and \R{summarize()} functions can be
used to run the corresponding visualizations and summaries for each
variable. See Figure~\ref{fig:example2} for the visualization output.

\begin{Schunk}
\begin{Sinput}
> visualize(toyData$var2)
> summarize(toyData$var2)
\end{Sinput}
\begin{Soutput}
     Feature                   Result    
[1,] "Variable type"           "numeric" 
[2,] "Number of missing obs."  "3 (20 %)"
[3,] "Number of unique values" "8"       
[4,] "Median"                  "4.5"     
[5,] "1st and 3rd quartiles"   "1.75; 6" 
[6,] "Min. and max."           "1; 999"  
\end{Soutput}
\end{Schunk}
% $

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-var2.pdf}
\end{center}
\label{fig:example2}
\caption{Output from running \R{visualize()} on the variable \texttt{var2} from the
\R{toyData} dataset.}
\end{figure}

As we saw with the \R{check()} function, the summary can be modified
by providing the relevant \R{XXXSummaries} arguments. Setting the
\R{numericSummaries} argument we can control the summary output by
providing a vector of functions to run for a particular summary. To
only get the minimum and maximum we set
\R{numericSummaries=c("minMax")}.

\begin{Schunk}
\begin{Sinput}
> summarize(toyData$var2, numericSummaries=c("minMax"))
\end{Sinput}
\begin{Soutput}
     Feature         Result  
[1,] "Min. and max." "1; 999"
\end{Soutput}
\end{Schunk}
% $

%\hl{
%\begin{itemize}
%\item Do an example with visualize() and summarize(), like the one with check(). Especially visualize and to doEval = T thing needs a bit of special attention.
% \item  Mention allCheckFunctions() etc. again here
% \item Mention check(), visualize() and summarize() modes for data.frames. Maybe also advice against it, at it will often produce a lot of information at once, and such large amounts of information really should be documented.
%\end{itemize}
%}

@@

\begin{table}[ht]
\scriptsize
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \R{summaryFunction} & \R{visualFunction} & \R{checkFunction} \\
\hline
Input (required) & \R{v} - a variable vector \newline \R{...}  &  \R{v} - a variable vector \newline \R{vnam} - the variable name (as character string) \newline \R{doEval} - a logical (\R{TRUE}/\R{FALSE}) controlling the output type of the function & \R{v} - a variable vector \newline \R{nMax} - an integer (or \R{Inf}), controlling how many problematic values are printed, if relevant \newline \R{...}  \\
Input (optional) &  \R{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \R{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g. a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries \newline \R{\$feature} - a label for the summary value (as character string) \newline \R{\$result} - the result of the summary (as character string) & A character string with \R{R} code for producing a plot. This code should be standalone, i.e. should include the data if necessary. & A list with entries \newline \R{\$problem} - a logical identifying whether an issue was found \newline \R{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \R{rmarkdown} \\
Output (recommended) & A \R{summaryResult} object (i.e. an attributed list with entries \R{\$feature}, \R{\$result} and \R{\$value}, the latter being the values from \R{\$result} in their original format). & \begin{tabular}{p{0.4\linewidth} | p{0.4\linewidth}} If \R{doEval} is \R{TRUE}: \newline A plot that will be opened by the graphic device in \R{R}. & If \R{doEval} is \R{FALSE}: \newline   A text string with \R{R} code, as described above. \end{tabular} & A \R{checkResult} object (an attributed list with entries \R{\$problem}, \R{\$message} and \R{\$problemValues}, the latter being either \R{NULL} or the problem causing values, as they were found in \R{v}, whichever is relevant. \\
Tools available for producing the function & \R{summaryResult()} & - & \R{messageGenerator()} \newline \R{checkResult()} 
\end{tabular}
\caption{Reference information for creating new functions to be used
  as part of the summarize, visualize, and check steps. The three
  columns correspond to each of the three function types.}
\label{table.functionTypes}
\end{table}

\pkg{cleanR} can be used as a user-friendly, self-contained package as
shown in the previous section. However, \pkg{cleanR} is fully
customizable and \R{clean()} is mainly a tool for formatting the
results from various checking-, summary- and visualization
functions. Thus, the actual work underlying a \pkg{cleanR} output file
can be anything --- depending on the arguments given to \R{clean()}
--- and user made functions are easily added to the
summarize/visualize/check-function arguments as discussed previously
(e.g., \R{factorSummaries}, or \R{numericChecks}). However, the
functions used in the summarize/visualize/check steps need to adhere
to a specific structure to be called from these three steps, and
therefore, we will now present how \R{summaryFunction}s,
\R{visualFunction}s and \R{checkFunction}s are made.
%\hl{Mention something about the interactive mode here as well?}

This section consists of two parts. First, we describe how the
\R{clean()} function can be extended by adding custom prechecks,
summaries, checks and visualizations. In order to do this, one needs
to create functions with a specific structure required by
\pkg{cleanR}. This can be done with different levels of strictness. If
the custom functions are only to be used part of the corresponding
arguments of \R{clean()} then only the input/output structures of the
functions need to be fixed.
